---
title: "Pruning fgsea results using the GO graph"
author: "Giuseppe D'Agostino"
date: "2/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

  
Load necessary libraries

```{r libraries}
library(GO.db)
library(fgsea)
library(BiocParallel)
library(msigdbr)
library(tidyverse)
library(ggplot2)
library(AnnotationDbi)
library(igraph)
library(Rgraphviz)
```

Read in the `DESeq2` results

```{r read_counts}
res.dds.lps.sh <- readRDS(file = "res_lps_sh.RDS")
```

Rank by the Wald test statistic

```{r rank}
lps.ranks.frame <- res.dds.lps.sh %>% 
  dplyr::select(entrezgene_id, stat) %>% 
  na.omit() %>% 
  distinct() %>% 
  group_by(entrezgene_id) %>% 
  summarize(stat=mean(stat))

lps.ranks <- deframe(lps.ranks.frame)
```

We get all the genesets in Gene Ontology (C5)

```{r get_genesets}
go.genesets <- msigdbr(species = "Mus musculus", category = "C5")
```

Prepare a list of genesets and convert IDs to Entrez

```{r wrangle_ids}
go.genesets = as.data.frame(go.genesets, stringsAsFactors = FALSE)
go.genesets$entrez_gene = as.character(go.genesets$entrez_gene)
go.list <- lapply(unique(go.genesets$gs_name), function(x) go.genesets[go.genesets$gs_name == x, "entrez_gene"])
names(go.list) = unique(go.genesets$gs_exact_source)  ##IMPORTANT: use gs_exact_source to retain the GO ID
```


Run `fgsea`

```{r run_fgsea}
fgsea.lps.go <- fgsea(pathways = go.list, 
                      stats = lps.ranks,
                      nperm = 10000,
                      BPPARAM = SnowParam())
```


Add names and subcategories to the fgsea results table

```{r}
goid_to_name <- as.data.frame(cbind(unique(go.genesets$gs_exact_source), unique(go.genesets$gs_name)))
goid_and_subcat <- as.data.frame(unique(go.genesets[,c("gs_subcat", "gs_exact_source")]))
rownames(goid_and_subcat) <- goid_and_subcat$gs_exact_source
colnames(goid_to_name) <- c("GO_ID", "name")
rownames(goid_to_name) <- goid_to_name$GO_ID
goid_to_name$subcat <- goid_and_subcat[rownames(goid_and_subcat), "gs_subcat"]

fgsea.lps.go$name <- goid_to_name[fgsea.lps.go$pathway, "name"]
fgsea.lps.go$subcat <- goid_to_name[fgsea.lps.go$pathway, "subcat"]
fgsea.lps.go$subcat <- gsub(fgsea.lps.go$subcat, pattern = "GO:", replacement = "")

fgsea.lps.go <- fgsea.lps.go[fgsea.lps.go$subcat != "HPO"] # we remove HP terms because they are not in the GO graph
fgsea.lps.go$padj <- p.adjust(fgsea.lps.go$pval, method = "fdr")
```


Here is the `fgsea` result pruning function: treats significant and non-significant categories separately. 

- Non-significant categories are only kept if they don't have children categories.

- Significant categories are only kept if they do not have enriched children categories

- Adjusted p-values are recalculated on the nominal p-values from the pruned categories.

```{r prune_function}
pruneGO_FGSEA <- function(fgsea_res, 
                         go_sub = c("BP", "MF", "CC"),
                         alpha = 0.05){
  
  fgsea_res <- fgsea_res[which(fgsea_res$subcat == go_sub),]

  if(go_sub == "BP") go_subcategory = GOBPCHILDREN 
    else  if(go_sub == "MF") go_subcategory = GOMFCHILDREN
    else  if(go_sub == "CC") go_subcategory = GOCCCHILDREN 
  
  go_children <- as.list(go_subcategory)
  go_haschildren <- go_children[!is.na(go_children)]
  
  fgsea_res$has_children <- sapply(fgsea_res$pathway, function(x) x %in% names(go_haschildren))

  fgsea_res_sig <- fgsea_res[fgsea_res$pval < alpha,]

  fgsea_res_sig$has_sig_children <- sapply(fgsea_res_sig$pathway, 
                                            function(x) any(go_haschildren[[x]] %in% fgsea_res_sig$pathway))
   
  fgsea_res_sig$keep <- (fgsea_res_sig$has_children == FALSE) | (fgsea_res_sig$has_children == TRUE & fgsea_res_sig$has_sig_children == FALSE) 
  
  fgsea_res_ns_childless <- fgsea_res[which(fgsea_res$pval >= alpha & !fgsea_res$has_children),]

  pruned <- rbind(fgsea_res_sig[fgsea_res_sig$keep,1:11], fgsea_res_ns_childless[,1:11])
  
  pruned$padj_2 <- p.adjust(pruned$pval, method = "fdr")
  
  pruned_len <- length(setdiff(pruned$pathway, fgsea_res_sig$pathway))
  message("Pruned ", pruned_len, " terms.")
  
  return(pruned)
  
 }
```

Prune separately by different subcategory (necessary to look for children in the right DAG)

```{r}
fgsea.lps.go.bp.pruned <- pruneGO_FGSEA(fgsea.lps.go, go_sub = "BP", alpha = 0.05)
fgsea.lps.go.mf.pruned <- pruneGO_FGSEA(fgsea.lps.go, go_sub = "MF", alpha = 0.05)
fgsea.lps.go.cc.pruned <- pruneGO_FGSEA(fgsea.lps.go, go_sub = "CC", alpha = 0.05)
```

Join all tables and perform (without overwriting) fdr correction

```{r join_tables}
fgsea.lps.go.all.pruned <- rbind(fgsea.lps.go.bp.pruned, fgsea.lps.go.mf.pruned, fgsea.lps.go.cc.pruned)
fgsea.lps.go.all.pruned$padj_3 <- p.adjust(fgsea.lps.go.all.pruned$pval, method = 'fdr')
```

Check nominal p value histogram as sanity check

```{r pval_hist}
hist(fgsea.lps.go.all.pruned$pval)
```

Select significant results

```{r select_sig}
fgsea.lps.go.all.pruned.sig <- fgsea.lps.go.all.pruned[fgsea.lps.go.all.pruned$padj_2 < 0.05,]
```

How many new significant terms did we gain? 

```{r terms_gained}
nrow(fgsea.lps.go.all.pruned.sig) -  length(which(fgsea.lps.go[fgsea.lps.go$pathway %in% fgsea.lps.go.all.pruned.sig$pathway,"padj"] < 0.05))

plot(x = -log10(fgsea.lps.go.all.pruned.sig$padj_2), y = -log10(as.data.frame(fgsea.lps.go)[fgsea.lps.go$pathway %in% fgsea.lps.go.all.pruned.sig$pathway,"padj"]), ylim = c(0, 2.3), xlim = c(0,2.3))
abline(h = -log10(0.05), col = 'red')
abline(v = -log10(0.05), col = 'red')
```

Interesting terms

```{r interesting_res_plot}
fgsea_interesting <- fgsea.lps.go.all.pruned.sig[grep("MITO|LIP|METAB|ENDOC|PHAG|GLU", fgsea.lps.go.all.pruned.sig$name),]

ggplot(fgsea_interesting, aes(x = reorder(name, NES), y = NES, col = subcat)) + 
  geom_point(aes(size = -log10(padj_2))) + 
  coord_flip() + 
  facet_wrap(~subcat) +
  theme_bw()

```

Top 30 terms (ranked by NES)

```{r top30}
fgsea_top30 <- fgsea.lps.go.all.pruned.sig[order(abs(fgsea.lps.go.all.pruned.sig$NES), decreasing = TRUE)[1:30],]

ggplot(fgsea_top30, aes(x = reorder(name, NES), y = NES, col = subcat )) + 
  geom_point(aes(size = -log10(padj_2))) + 
  coord_flip() + 
  facet_wrap(~subcat) +
  theme_bw()

```

Top 30 in the non-pruned table:

```{r top30_nonpruned}
fgsea.lps.go$padj_2 <- p.adjust(fgsea.lps.go$pval, method = "fdr")
fgsea.lps.go.sig <- fgsea.lps.go[fgsea.lps.go$padj_2 < 0.05,]
fgsea_top30_nonpruned <- fgsea.lps.go.sig[order(abs(fgsea.lps.go.sig$NES), decreasing = TRUE)[1:30],]

ggplot(fgsea_top30_nonpruned, aes(x = reorder(name, NES), y = NES, col = subcat )) + 
  geom_point(aes(size = -log10(padj_2))) + 
  coord_flip() + 
  facet_wrap(~subcat) +
  theme_bw()

```


Did the pruning algorithm work?

```{r, check_graphs}
graph.par(list(nodes=list(fontsize=40)))

fgsea.lps.go.sig.nominal <- fgsea.lps.go[fgsea.lps.go$pval < 0.05,]
fgsea.lps.go.mf.pruned.sig <- fgsea.lps.go.mf.pruned[fgsea.lps.go.mf.pruned$padj_2 < 0.05,]

mfgraph <- igraph::graph_from_graphnel(makeGOGraph("mf")) 
mfgraph_induced <- igraph::subgraph(mfgraph, v = which(names(V(mfgraph)) %in% fgsea.lps.go.sig.nominal$pathway))
mfgraphnel <- igraph::igraph.to.graphNEL(mfgraph_induced)
mfgraphnel <- layoutGraph(mfgraphnel)
nodeRenderInfo(mfgraphnel)$fill[fgsea.lps.go.mf.pruned$pathway] <- 'red' 
nodeRenderInfo(mfgraphnel)$col[fgsea.lps.go.mf.pruned$pathway] <- 'red' 
nodeRenderInfo(mfgraphnel)$fill[fgsea.lps.go.mf.pruned.sig$pathway] <- 'blue' 
nodeRenderInfo(mfgraphnel)$col[fgsea.lps.go.mf.pruned.sig$pathway] <- 'blue' 
renderGraph(mfgraphnel)

fgsea.lps.go.bp.pruned.sig <- fgsea.lps.go.bp.pruned[fgsea.lps.go.bp.pruned$padj_2 < 0.05,]

bpgraph <- igraph::graph_from_graphnel(makeGOGraph("bp"))
bpgraph_induced <- igraph::subgraph(bpgraph, v = which(names(V(bpgraph)) %in% fgsea.lps.go.sig.nominal$pathway))
bpgraphnel <- igraph::igraph.to.graphNEL(bpgraph_induced)
bpgraphnel <- layoutGraph(bpgraphnel)
nodeRenderInfo(bpgraphnel)$fill[fgsea.lps.go.bp.pruned$pathway] <- 'red' 
nodeRenderInfo(bpgraphnel)$col[fgsea.lps.go.bp.pruned$pathway] <- 'red' 
nodeRenderInfo(bpgraphnel)$fill[fgsea.lps.go.bp.pruned.sig$pathway] <- 'blue' 
nodeRenderInfo(bpgraphnel)$col[fgsea.lps.go.bp.pruned.sig$pathway] <- 'blue' 
renderGraph(bpgraphnel)


fgsea.lps.go.cc.pruned.sig <- fgsea.lps.go.cc.pruned[fgsea.lps.go.cc.pruned$padj_2 < 0.05,]

ccgraph <- igraph::graph_from_graphnel(makeGOGraph("cc"))
ccgraph_induced <- igraph::subgraph(ccgraph, v = which(names(V(ccgraph)) %in% fgsea.lps.go.sig.nominal$pathway))
ccgraphnel <- igraph::igraph.to.graphNEL(ccgraph_induced)
ccgraphnel <- layoutGraph(ccgraphnel)
nodeRenderInfo(ccgraphnel)$fill[fgsea.lps.go.cc.pruned$pathway] <- 'red' 
nodeRenderInfo(ccgraphnel)$col[fgsea.lps.go.cc.pruned$pathway] <- 'red' 
nodeRenderInfo(ccgraphnel)$fill[fgsea.lps.go.cc.pruned.sig$pathway] <- 'blue' 
nodeRenderInfo(ccgraphnel)$col[fgsea.lps.go.cc.pruned.sig$pathway] <- 'blue'  
renderGraph(ccgraphnel)
```
